PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2023-02-23 reviewer_response.tex

\reviewer{R2.6}{Line 183 - I really do not understand what is going on with your alpha,beta being set to zero. Are you saying if you stop penalising merges for having too many SNPs, then you will be able to merge more, and this will stop you from havign a 10\% divergence limit? It will also mean you make more false merges, right?}
% \Marco{setting $\beta$ to zero will indeed remove the 10\% divergence limit set by the pseudo-energy. There is another limit that depends on the aligner sensitivity (20 to 30\%, depending on the aligner). This limit remains, and in this sense no ``false merges'' are performed? Not sure I interpret this correctly.}\\
% \Liam{I think your interpretation is right. Important to clarify that there are those two limits and that $\beta$ only controls the divergence limit \textit{within} what one can actually align.}
\response{There are two factors limiting the divergence of merges that are performed. The first factor is the aligner used. Minimap2 is in general able to find matches up to roughly 20\% sequence divergence, while mmseqs2 can reach divergence up to roughly 30\%. Only homology matches of up to this divergence will be suggested by the aligners. On top of this, only matches with negative pseudo-energy are turned into merges. For any given value of $\beta$, this sets an additional limit to the divergence, which must be less than $1/\beta$.\\
      This is now made more explicit in SI section I (PanGraph parameters effect).\\
      When setting $\beta = 0$ only the intrinsic limit on divergence posed by the aligner sensitivity is left. In this sense more merges are allowed, but all mergess will have some level of homology (maximum 20\% or 30\% divergence, depending on the aligner).}

\reviewer{R2.7}{Why does the L50 peak go up for null-scores for mmseqs for Mtb, Pm, Kpn and Ec (Figure 4E, purple with lines) but down for helicobacter?}
% \Marco{the effect of merging more blocks on L50 is not trivial. It depends on whether the merging occurs between long or short blocks, and how much fragmentation this causes. HP is a special case in that its diverge is on the threshold of what minimap2 can handle. Using mmseqs2 results in a pangenome that is half the size compared to standard parameters (fig. 4G). This is an impressive amount of additional compression. At the same time the fraction of core pangenome more than doubles (4H). The total number of blocks decreases to roughly half (4D) while N50 increases (4F). For me this points to the fact that in the case of HP merging will split long blocks and create more fragmentation. For the reviewer I could produce the block frequency and lenght distribution for the two cases and comparing them would give an answer. I could do the same for EC or KP for comparison.}\\
% \Liam{makes sense.}
\response{L50 measures the minimum number of pancontigs needed to cover at least 50\% of the pangenome graph sequence. This number is closely related to the total number of pancontigs. In general the use of null energy parameters result in more fragmented graphs, with a higher number of pancontigs (cf. fig.~4D in SI). Hp is an exception and has the opposite trend, both for the number of pancontig and for the value of L50. For this species the use of null energy parameters results in less fragmentation of the graph. This is compatible with the observation that the average divergence of this dataset is on the limit of what PanGraph can handle. Unless a highly sensitive aligner is used and the divergence barrier in pseudo-energy is removed, some highly-diverged homologous sequences remain split in different pancontigs. Conversely, the choice of these parameters generates a graph with less pancontigs, a lower L50, a larger fraction of core pangenome, an higher median length of pancontigs (N50).\\
      For a visual verification of this interpretation we display in fig.~\ref{fig:kp-vs-hp} how the distributions of pancontig lenght and frequency change with the use of different alignment kernels and pseudo-energy parameters. We compare the changes observed on the Hp dataset to the ones in Kp, used as a reference. In the Kp dataset increasing the aligner sensitivity results in shorter pancontigs (lower-right panel) and a small increase in core genome fraction. For Hp instead a more sensitive aligner results in the loss of small and intermediate size pancontigs, but not of long ones. It also results in a loss of low-frequency pancontigs, indicating that these had an homologous counterpart that could not be merged due to high-divergence, and the core pangenome fraction increases considerably.
      % TODO: add mention in SI
      % QUESTION: this is more or less explicit in SI? Should I expand more?
}

\begin{figure}[htb]
      \includegraphics[width=.7\textwidth]{figs_response/hp_vs_kp.pdf}
      \caption{frequency and length distribution (respectively left and right panels) of pancontigs for the \textit{Helicobacter pylori} and \textit{Klebsiella pneumoniae} datasets (top and bottom). For each panel we report the results for pangenome graphs built using minimap2 aligner with standard parameter value and asm20 sensitivity (blue) and with mmseqs2 and null pseudo-energy parameters (orange). Distributions are weighed by pancontig length, so that the area under the curve represents the total pangenome graph size.}
      \label{fig:kp-vs-hp}
\end{figure}

@@

\reviewer{R2.8}{Just to make sure i understand - is it possible for sequence to be ``left out'' in between pancontigs?}
\Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}\\
\Liam{if they possibly misunderstood, suggests that some statement that all sequence ends up in a pancontig should be added to manuscript. I suspect they are thinking of case of short ($<$100bp) stretches `in between' pancontigs, which (if I understand correctly) will be incorporated as insertions at the end of the consensus alignment of the adjacent pancontig.}

\reviewer{R2.9}{About data accessibility
      \begin{enumerate}
            \item i don't see how to get the list of accessions for section IIB? Saying they were used in the Ding et al paper is not quite enough for me to be sure which ones you used.
            \item for the graph marginalisation section, could the accessions of the strains used be in a supplementary file as well as on github? Github repos are great, but they can be deleted.
      \end{enumerate}}
% \Marco{on the repo we have \href{https://github.com/neherlab/pangraph/blob/master/script/config/accnums.json}{this file} for the list of all accession numbers we use, and \href{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}{this file} for the marginalization pairs. I could create some excel spreadsheet with accession numbers that we could add to the submission?}\\
% \Liam{for sure that's all they want (most biologists don't like json!)}
\responsedraft{We now included in the submission two excel files. ``accession\_numbers.xlsx'' contains five sheets, one per species considered, each with a list of the accession numbers of isolates used in our datasets. ``marginalize\_pairs.xlsx'' contains instead a list of all the pairs used in the graph marginalization section.
      % TODO: archive repository state.
}

### 2023-02-23 main.tex
update

We then explored how the properties of the pangraph scale with the dataset size. To do so, we build pangenome graphs from randomized subsets of isolates of increasing size from the \textit{E.~coli} dataset. Results are reported in Fig.~\ref{fig:panx-size}.\\
The number and size of pancontigs scale sub-linearly with the number of isolates, with more than 50\% of the pangenome being included in the 10\% longest pancontigs. Core pancontigs having higher-than-average length. This is expected given their strong synteny and the fact that they present less structural variation across genomes compared to accessory genes. Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added. The size of the core genome does not decrease significantly with the addition of new strains, while the total pangenome size remains orders of magnitude smaller than the total size of all genomes included in the graph.

@@

\begin{figure*}[htb]
    \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
    \caption{{\bf Test of graph marginalization}.
        \textbf{A}: we built a pangenome graph from 50 randomly chosen strains from the \textit{K.~pneumoniae} dataset. We then randomly picked 50 pairs of strains. For each pair we compared the pangenome graph obtained by marginalizing the complete graph on the pair of strain, and the one obtained by building a new graph for the pair (top). The comparison is done by considering that each graph partitions a genome in shared and private segments. By combining the partitions generated by the marginalized and pairwise graphs we categorize segments in three categories, depending on whether the two partitions agree or not, and if they agree depending on whether segments are shared or private. All graphs were built using \textit{minimap2} alignment kernel with \textit{asm20} option and default value for the energy parameters.
        \textbf{B}: distribution of the average fraction of the genome covered by segments of each category, over the 50 pairs considered (2 entries per pair). The last line represents the distribution of shared sequence, approximated using the fraction of shared k-mers corrected using sequence divergence as described in the main text. Next to each distribution we report its mean and standard deviation.
        \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair). Mean and standard deviation are reported.
    }
    \label{fig:marginalization}
\end{figure*}

>>>>

We then explored how the properties of the pangraph scale with the dataset size. To do so, we build pangenome graphs from randomized subsets of isolates of increasing size from the \textit{E.~coli} dataset. Results are reported in Fig.~\ref{fig:panx-size}.\\
The number and size of pancontigs scale sub-linearly with the number of isolates, with more than 50\% of the pangenome being included in the 10\% longest pancontigs. \oldnew{, and core pancontigs having higher-than-average size}{Core pancontigs have higher-than-average length. This is expected given their strong synteny and the fact that they present less structural variation across genomes compared to accessory genes.} Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added. The size of the core genome does not decrease significantly with the addition of new strains, while the total pangenome size remains orders of magnitude smaller than the total size of all genomes included in the graph.

@@

\begin{figure*}[htb]
    \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
    \caption{{\bf Test of graph marginalization}.
        \textbf{A}: we built a pangenome graph from 50 randomly chosen strains from the \textit{K.~pneumoniae} dataset. We then randomly picked 50 pairs of strains. For each pair we compared the pangenome graph obtained by marginalizing the complete graph on the pair of strain, and the one obtained by building a new graph for the pair (top). The comparison is done by considering that each graph partitions a genome in shared and private segments. By combining the partitions generated by the marginalized and pairwise graphs we categorize segments in three categories, depending on whether the two partitions agree or not, and if they agree depending on whether segments are shared or private. All graphs were built using \textit{minimap2} alignment kernel with \textit{asm20} option and default value for the energy parameters.
        \textbf{B}: distribution of the average fraction of the genome covered by segments of each category, over the 50 pairs considered (2 entries per pair). The last line represents the distribution of shared sequence, approximated using the fraction of shared k-mers corrected using sequence divergence as described in the main text. \oldnew{}{Next to each distribution we report its mean and standard deviation.}
        \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair). Mean and standard deviation are reported.
    }
    \label{fig:marginalization}
\end{figure*}

### 2023-02-23 supplementary.tex
update

As a first test we benchmark the performances and accuracy of \textit{PanGraph} on synthetic data, obtained by simulating the evolution of a population in which individuals accumulate mutations, can loose part of the sequence from deletions, or gain new genetic material via Horizontal Sequence Transfer (HST) from other members of the population.\\
In the simulation a population of size $N$ is evolved for $T$ generations using a Wright-Fisher model \cite{hudson2002generating}. Each individual in the population has a circular genome of initial size $L$. At each generation single-nucleotide mutations occur at a rate $\mu$ per position. Deletions can occur at a rate $d$ per genome per generation. When a deletion is suggested on isolate $n$, having a genome length $L_n$, a new desired length $L'_n$ is extracted from a Gaussian distribution with mean $L$ and variance $\sigma^2$. If $\Delta = L'_n - L_n < 0$ then a random chunk of length $|\Delta|$ is removed from the genome, reducing it to size $L'_n$. If $\Delta > 0$ no deletion is performed. This ensures that the length distribution of genomes in the population remains close to the desired length $L$, with $\sigma^2$ being a proxy for the variance.
HST occurs at a rate $h$ per genome per generation. Similarly to deletions, when a HST event is suggested a new desired length $L'_n$ is extracted from the same Gaussian distribution, and this time the event is performed only if $\Delta = L'_n - L_n > 0$. In this case a random chunk of sequence of length $\Delta$ is extracted from a random individual in the parent population (possibly the parent of isolate $n$ itself) and inserted in a random position of the genome of isolate $n$. All parts of the sequence have equal probability of being transferred, irrespective of homology.
Finally, inversions occur at a rate $i$ per genome per generation. When an inversion is proposed a random length $\Delta$ is chosen in the same way, and a random chunk of sequence of this length is inverted. In the simulation we do not implement a notion of genes.
Standard values of the simulation parameters are reported in Table~\ref{table:sim-params}.

>>>>

As a first test we benchmark the performances and accuracy of \textit{PanGraph} on synthetic data, obtained by simulating the evolution of a population in which individuals accumulate mutations, can loose part of the sequence from deletions, or gain new genetic material via Horizontal \oldnew{Gene}{Sequence} Transfer (HST) from other members of the population.\\
In the simulation a population of size $N$ is evolved for $T$ generations using a Wright-Fisher model \cite{hudson2002generating}. Each individual in the population has a circular genome of initial size $L$. At each generation single-nucleotide mutations occur at a rate $\mu$ per position. Deletions can occur at a rate $d$ per genome per generation. When a deletion is suggested on isolate $n$, having a genome length $L_n$, a new desired length $L'_n$ is extracted from a Gaussian distribution with mean $L$ and variance $\sigma^2$. If $\Delta = L'_n - L_n < 0$ then a random chunk of length $|\Delta|$ is removed from the genome, reducing it to size $L'_n$. If $\Delta > 0$ no deletion is performed. This ensures that the length distribution of genomes in the population remains close to the desired length $L$, with $\sigma^2$ being a proxy for the variance.
HST occurs at a rate $h$ per genome per generation. Similarly to deletions, when a HST event is suggested a new desired length $L'_n$ is extracted from the same Gaussian distribution, and this time the event is performed only if $\Delta = L'_n - L_n > 0$. In this case a random chunk of sequence of length $\Delta$ is extracted from a random individual in the parent population (possibly the parent of isolate $n$ itself) and inserted in a random position of the genome of isolate $n$. \oldnew{}{All parts of the sequence have equal probability of being transferred, irrespective of homology.}
Finally, inversions occur at a rate $i$ per genome per generation. When an inversion is proposed a random length $\Delta$ is chosen in the same way, and a random chunk of sequence of this length is inverted. \oldnew{}{In the simulation we do not implement a notion of genes}.
Standard values of the simulation parameters are reported in Table~\ref{table:sim-params}.
